Limb-darkening and exoplanets II: Choosing the Best Law for Minimum Bias Retrieval of Transit Parameters

### 2015-11-13 which_law_2_use.tex
Added small changes to the final revised version.

\title[Choosing the best law for minimum bias retrieval of transit parameters]
{Limb-darkening and exoplanets II: Choosing the Best Law for Minimum Bias Retrieval of Transit Parameters}
\author[Espinoza \& Jord\'an ]{
N\'estor Espinoza$^{1,2}$\thanks{E-mail:nespino@astro.puc.cl}, 
Andr\'es Jord\'an$^{1,2}$\thanks{E-mail:ajordan@astro.puc.cl}\\ 
$^{1}$Instituto de Astrof\'isica, Pontificia Universidad Cat\'olica de Chile, Vicu\~na Mackenna 
4860, Santiago, Chile
$^{2}$Millennium Institute of Astrophysics, Santiago, Chile\\
}
\begin{document}

@@

In a recent study \citep[][hereon EJ15]{ej2015}, we showed that the accuracy is actually catching up with the precision 
of these measurements due to our imperfect understanding of the limb-darkening effect. In particular, we showed that there are important 
biases both when fixing and fitting for the limb-darkening coefficients in the light curve fitting process. The biases when fitting for the limb darkening coefficients arise from the fact 
that the popular and widely used quadratic law is unable to model the complex intensity profile of real stars,  and the biases when fixing the coefficients arise for the
same reason plus the facts that different methods of fitting the model intensity profiles (transit light curve versus intensity profile fitting) give rise to different limb-darkening coefficients and 
the fact that we have an imperfect knowledge of the real intensity profiles of stars. We showed that fixing the limb-darkening coefficients is 
the worst option if one aims to obtain precise \textit{and} accurate transit parameters, because the biases arise from three different sources, 
with the last one (the fact that we do not have a perfect modelling of the stellar intensity profiles of real stars) having an unknown but possibly 
large impact.

Given the above, unless the data quality is really poor (and one is willing to trade bias for variance), there is actually no good reason to fix the limb-darkening coefficients in the transit fitting process (assuming computational resources are not an issue). However, although fitting the 
limb-darkening coefficients seems to be a good solution to the accuracy problem, there remains the issue of the low flexibility of the quadratic limb-darkening 
law which can cause important biases on the retrieved transit parameters as shown in EJ15. These biases can be a large as $\sim 1\%$ for the planet-to-star 
radius ratio $R_p/R_*$, $\sim 2\%$ for the scaled semi-major axis, $a/R_*$, and $2\%$ for the inclination, where $R_p$ is the planetary radius, $R_*$ 
the stellar radius and $a$ the semi-major axis. The importance for missions like {\em Kepler} and for future missions like the Transiting Exoplanet Survey 
Satellite \cite[TESS,][]{ricker2014} is evident when looking at the most recent results from the {\em Kepler} mission: if we focus on the planet-to-star radius ratio measurements 
alone, which according to EJ15 have an accuracy bias on the order of $\sim 0.2\%$, a query to the Nasa Exoplanet Archive\footnote{\url{http://exoplanetarchive.ipac.caltech.edu/}; query done on 29/09/2015.} shows that out of 3063 planetary candidates, 933 ($26\%$) have precisions better than 
$0.2\%$, and out of 1001 {\em Kepler} confirmed exoplanets, 463 (46\%) do too. This means that at least\footnote{It is important to note that this is an 
underestimate of the number of planet candidates with estimation errors, due to the fact that the Kepler pipeline makes use of fixed limb-darkening 
coefficients using the quadratic law \citep{rowe2015}, which as explained here and estimated in EJ15, gives rise to more severe biases than the one used for this 
estimation} $26\%$ of the planet candidates (from which, e.g., population studies, which rely on averaging out \textit{random} and not \textit{systematic} 
uncertainties like the ones introduced by limb-darkening) and almost half of the confirmed exoplanets (from which, e.g., characterization studies are 
based on) have significant systematic errors.

An obvious solution to the issues mentioned above if one decides to fit the limb-darkening coefficients is to try to use alternative laws to describe 
the intensity profile of stars. Although the non-linear law proposed by \cite{claret2000} seems to be the most flexible, the fact that it has four free parameters 
does not make it a very attractive. Several laws with fewer parameters have been proposed. Two-parameter laws are the exponential proposed by \cite{claret2003}, the logarithmic proposed by \cite{klinglesmith1970} and the square-root proposed 
by \cite{diazgimenez1992}. In addition, a very flexible three-parameter law was also proposed by \cite{sing2009}. These laws with fewer parameters are potentially 
very attractive for transit fitting purposes, due to their flexibility at following different intensity profiles  and their low number of parameters \citep[see, e.g., ][for a comparison between the goodness of 
fit to ATLAS model atmospheres of the mentioned two-parameter laws which outperform the linear and quadratic laws in terms of following the intensity 
profiles]{howarth2011}.

Despite the attractive nature of the above mentioned limb-darkening laws, testing their performance at retrieving transit parameters from transit 
lightcurves was a problem until recently for two reasons. First, there was no publicly available code capable of efficiently generating fast and accurate transit lightcurves using all of these non-standard laws. However, recently \cite{batman2015} published an 
algorithm that does exactly this called \texttt{batman}, enabling the generation of transit lightcurves very efficiently with any arbitrary limb-darkening law. 
The second problem was that it was not clear how to sample limb-darkening coefficients in an informative (i.e., sampling all the physically possible parameter 
space) and efficient way for all of these laws. Recently, \cite{kipping2015} presented an algorithm to sample parameters from the three-parameter law by imposing physically motivated constraints on the intensity profiles which derived in constrains on the limb-darkening coefficients being fitted. cite{kipping2013}, using the same principles, derived an algorithmic way of 
doing this for two-parameter laws. He found that his algorithm was not applicable to the exponential law and thus no constrains for this law could 
be derived. We note that the form of the logarithmic law used in \cite{kipping2013} differs from the standard one proposed by 
\cite{klinglesmith1970}. Thus, in order to  efficiently sample coefficients from the latter limb-darkening law one has to derive an efficient  informative sampling scheme for it.

@@

We note that typical values obtained by fitting the exponential law to model atmospheres give $e_2<0$, which means that 
most of the profiles fitted to model atmospheres do not have an everywhere positive intensity profile. To illustrate 
this, in Figure~\ref{exponential_ldcs} we plot all the coefficients fitted using the exponential law following the methods 
of EJ15 using the {\em Kepler} bandpass for model atmospheres with $T_\textnormal{eff}<9000$ K from  
the ATLAS models. In order to 
see at which value of $\mu$ the profile starts to be negative, we note that this happens as $\mu \to 0^+$. Here 
$1/(1-e^\mu) \approx -1/\mu$, which gives
\begin{eqnarray*}
I(\mu) &\approx& 1-e_1(1-\mu)+e_2/\mu.
\end{eqnarray*}
The intensity profile then, touches $I(\mu)=0$ around
\begin{eqnarray*}
\mu_0 \approx \frac{e_1-1 \pm \sqrt{(1-e_1)^2 - 4e_1e_2}}{2e_1}
\end{eqnarray*}
As shown Figure~\ref{exponential_ldcs}, $e_1\sim 0.6$, while $e_2 \sim - 2\times 10^{-3}$, which gives $\mu_0 \sim 0.005$, 
so the law starts giving negative values very close to (but not at) the limb, as we predicted. We note that typical transit observations are capable 
of sampling these points with high cadence observations as the ones done by the {\em Kepler} mission and the ones that will 
be done with the future TESS mission \citep{ricker2014}, which will have an almost certain chance of sampling points with $\mu\approx 1$. 
Furthermore, as the generation of transit lightcurves for laws like the exponential requiere either numerical integration or the usage 
of  approximations that usually sample more points close to the limb \citep[which is the case for e.g., the \texttt{batman} code,][]{batman2015}, this law is doomed to produce numerical errors in transit lightcurves due to this fact. Because of these reasons, we advise not to use this law.

@@

Having dealt with the sampling of limb-darkening coefficients from the 
logarithmic and exponential laws, we now turn to the study of the 
biases introduced on retrieved transit parameters in the transit fitting process introduced by using different limb-darkening laws. We perform a similar study as the one done in EJ15, but now 
(1) using the \texttt{batman} transit code described in \cite{batman2015}, (2) using three different two-parameter 
laws: the quadratic, logarithmic and square-root laws, and the three-parameter law and (3) smaller steps in 
the planet-to-star radius ratio, $p=R_p/R_*$, in order to illustrate the evolution of the biases with this parameter 
in a cleaner way: small ratios ($p=0.01$), medium ratios ($p=0.07$) and large ones ($p=0.13$).

>>>>

\title[Choosing the best law for minimum bias retrieval of transit parameters]
{Limb-darkening and exoplanets II: Choosing the Best Law for Minimum Bias Retrieval of Transit Parameters}
\author[Espinoza \& Jord\'an ]{
N\'estor Espinoza$^{1,2}$\thanks{E-mail:nespino@astro.puc.cl}, 
Andr\'es Jord\'an$^{1,2}$\thanks{E-mail:ajordan@astro.puc.cl}\\ 
$^{1}$Instituto de Astrof\'isica, Pontificia Universidad Cat\'olica de Chile, Vicu\~na Mackenna 
4860, Santiago, Chile\\
$^{2}$Millennium Institute of Astrophysics, Santiago, Chile\\
}
\begin{document}

@@

In a recent study \citep[][hereon EJ15]{ej2015}, we showed that the accuracy is actually catching up with the precision 
of these measurements due to our imperfect understanding of the limb-darkening effect. In particular, we showed that there are important 
biases both when fixing and fitting for the limb-darkening coefficients in the light curve fitting process. The biases when fitting for the limb darkening coefficients arise from the fact 
that the popular and widely used quadratic law is unable to model the complex intensity profile of real stars,  and the biases when fixing the coefficients arise for the
same reason plus the facts that different methods of fitting the model intensity profiles (including transit light curve versus intensity profile fitting) give rise to different limb-darkening coefficients and 
the fact that we have an imperfect knowledge of the real intensity profiles of stars. We showed that fixing the limb-darkening coefficients is 
the worst option if one aims to obtain precise \textit{and} accurate transit parameters, because the biases arise from three different sources, 
with the last one (the fact that we do not have a perfect modelling of the stellar intensity profiles of real stars) having an unknown but possibly 
large impact.

Given the above, unless the data quality is really poor (and one is willing to trade bias for variance), there is actually no good reason to fix the limb-darkening coefficients in the transit fitting process (assuming computational resources are not an issue). However, although fitting the 
limb-darkening coefficients seems to be a good solution to the accuracy problem, there remains the issue of the low flexibility of the quadratic limb-darkening 
law which can cause important biases on the retrieved transit parameters as shown in EJ15. These biases can be as large as $\sim 1\%$ for the planet-to-star 
radius ratio $R_p/R_*$, $\sim 2\%$ for the scaled semi-major axis, $a/R_*$, and $2\%$ for the inclination, where $R_p$ is the planetary radius, $R_*$ 
the stellar radius and $a$ the semi-major axis. The importance for missions like {\em Kepler} and for future missions like the Transiting Exoplanet Survey 
Satellite \cite[TESS,][]{ricker2014} is evident when looking at the most recent results from the {\em Kepler} mission: if we focus on the planet-to-star radius ratio measurements 
alone, which according to EJ15 have an accuracy bias on the order of $\sim 0.2\%$, a query to the Nasa Exoplanet Archive\footnote{\url{http://exoplanetarchive.ipac.caltech.edu/}; query done on 29/09/2015.} shows that out of 3063 planetary candidates, 933 ($26\%$) have precisions better than 
$0.2\%$, and out of 1001 {\em Kepler} confirmed exoplanets, 463 (46\%) do too. This means that at least\footnote{It is important to note that this is an 
underestimate of the number of planet candidates with estimation errors, due to the fact that the Kepler pipeline makes use of fixed limb-darkening 
coefficients using the quadratic law \citep{rowe2015}, which as explained here and estimated in EJ15, gives rise to more severe biases than the one used for this 
estimation} $26\%$ of the planet candidates (from which, e.g., population studies, which rely on averaging out \textit{random} and not \textit{systematic} 
uncertainties like the ones introduced by limb-darkening) and almost half of the confirmed exoplanets (from which, e.g., characterization studies are 
based on) have significant systematic errors.

An obvious solution to the issues mentioned above if one decides to fit the limb-darkening coefficients is to try to use alternative laws to describe 
the intensity profile of stars. Although the non-linear law proposed by \cite{claret2000} seems to be the most flexible, the fact that it has four free parameters 
does not make it a very attractive one. Several laws with fewer parameters have been proposed. Two-parameter laws are the exponential proposed by \cite{claret2003}, the logarithmic proposed by \cite{klinglesmith1970} and the square-root proposed 
by \cite{diazgimenez1992}. In addition, a very flexible three-parameter law was also proposed by \cite{sing2009}. These laws with fewer parameters are potentially 
very attractive for transit fitting purposes, due to their flexibility at following different intensity profiles  and their low number of parameters \citep[see, e.g., ][for a comparison between the goodness of 
fit to ATLAS model atmospheres of the mentioned two-parameter laws which outperform the linear and quadratic laws in terms of following the intensity 
profiles]{howarth2011}.

Despite the attractive nature of the above mentioned limb-darkening laws, testing their performance at retrieving transit parameters from transit 
lightcurves was a problem until recently for two reasons. First, there was no publicly available code capable of efficiently generating fast and accurate transit lightcurves using all of these non-standard laws. However, recently \cite{batman2015} published an 
algorithm that does exactly this called \texttt{batman}, enabling the generation of transit lightcurves very efficiently with any arbitrary limb-darkening law. 
The second problem was that it was not clear how to sample limb-darkening coefficients in an informative (i.e., sampling all the physically possible parameter 
space) and efficient way for all of these laws. Recently, \cite{kipping2015} presented an algorithm to sample parameters from the three-parameter law by imposing physically motivated constraints on the intensity profiles which derived in constrains on the limb-darkening coefficients being fitted. \cite{kipping2013}, using the same principles, derived an algorithmic way of 
doing this for two-parameter laws. He found that his algorithm was not applicable to the exponential law and thus no constrains for this law could 
be derived. We note that the form of the logarithmic law used in \cite{kipping2013} differs from the standard one proposed by 
\cite{klinglesmith1970}. Thus, in order to  efficiently sample coefficients from the latter limb-darkening law one has to derive an efficient  informative sampling scheme for it.

@@

We note that typical values obtained by fitting the exponential law to model atmospheres give $e_2<0$, which means that 
most of the profiles fitted to model atmospheres do not have an everywhere positive intensity profile. To illustrate 
this, in Figure~\ref{exponential_ldcs} we plot all the coefficients fitted using the exponential law following the methods 
of EJ15 using the {\em Kepler} bandpass for model atmospheres with $T_\textnormal{eff}<9000$ K from  
the ATLAS models. In order to 
see at which value of $\mu$ the profile starts to be negative, we note that this happens as $\mu \to 0^+$. Here 
$1/(1-e^\mu) \approx -1/\mu$, which gives
\begin{eqnarray*}
I(\mu) &\approx& 1-e_1(1-\mu)+e_2/\mu.
\end{eqnarray*}
The intensity profile then, touches $I(\mu)=0$ around
\begin{eqnarray*}
\mu_0 \approx \frac{e_1-1 \pm \sqrt{(1-e_1)^2 - 4e_1e_2}}{2e_1}
\end{eqnarray*}
As shown Figure~\ref{exponential_ldcs}, $e_1\sim 0.6$, while $e_2 \sim - 2\times 10^{-3}$, which gives $\mu_0 \sim 0.005$, 
so the law starts giving negative values very close to (but not at) the limb, as we predicted. We note that typical transit observations are capable 
of sampling these points with high cadence observations as the ones done by the {\em Kepler} mission and the ones that will 
be done with the future TESS mission \citep{ricker2014}, which will have an almost certain chance of sampling points with $\mu\lesssim \mu_0$. 
Furthermore, as the generation of transit lightcurves for laws like the exponential requiere either numerical integration or the usage 
of  approximations that usually sample more points close to the limb \citep[which is the case for e.g., the \texttt{batman} code,][]{batman2015}, this law is doomed to produce numerical errors in transit lightcurves due to this fact. Because of these reasons, we advise not to use this law.

@@

Having dealt with the sampling of limb-darkening coefficients from the 
logarithmic and exponential laws, we now turn to the study of the 
biases introduced on retrieved transit parameters in the transit fitting process introduced by using different limb-darkening laws. We perform a similar study as the one done in EJ15, but now 
(1) using the \texttt{batman} transit code described in \cite{batman2015}, (2) using three different two-parameter 
laws: the quadratic, logarithmic and square-root laws, and the three-parameter law and (3) smaller number of steps in 
the planet-to-star radius ratio, $p=R_p/R_*$, in order to illustrate the evolution of the biases with this parameter 
in a cleaner way: small ratios ($p=0.01$), medium ratios ($p=0.07$) and large ones ($p=0.13$).
